---
title: "Hands-on: Workflow Management"
format:
  html:
    toc-depth: 4
# The solution blocks below contain ```{r} / ```{python} fences; force the
# plain markdown engine so Quarto does not try to bind knitr or jupyter.
engine: markdown
execute:
  eval: false
---

# Workflow Demo: One Pipeline, Two Engines

In this hands-on you will run the same 3-step analysis pipeline using two different workflow engines **Nextflow** and **Snakemake** and observe how each one manages execution order, caching, and re-runs.

The pipeline is identical in both cases:

```
example_data.csv
      │
      ▼
 1_derive.py  (Python)  →  1_derived.csv   # mean-centre the values
                                │
                                ▼
                       2_summary.R  (R)  →  2_summary.txt   # summarise the centred values
                                                  │
                                │
                                ▼
                 3_polars_process.py  (Python/polars)  →  3_processed.csv   # add a polars-derived feature
                                                  │
                            ┌─────────────────────┘
                            ▼
                      4_report.qmd  (Quarto)  →  4_report.html   # collect outputs into a report
```

You do **not** need to understand the Python or R code in detail. The focus is on the workflow layer around them.

For the workshop, step 3 intentionally introduces an extra dependency (`polars`) that may not be installed by default.

---

## Nextflow

**Directory:** `demo/nextflow/`

The pipeline is defined in `main.nf`. Each step is a `process` block with explicit `input:` and `output:` declarations. Nextflow runs every process in an isolated subdirectory under `work/` and copies final outputs to `results/` via `publishDir`.

### Running

```bash
# Build the image and start the container (first run only)
docker compose up -d

# Open a shell inside the container
docker compose exec nextflow bash

# Run the pipeline
nextflow run main.nf
```

Results are written to `results/4_report.html`.

### Re-running after a change

Nextflow caches every task by content hash in `work/`. Pass `-resume` to skip unchanged steps:

```bash
nextflow run main.nf -resume
```

If you edit `2_summary.R`, only `R_SUMMARY` and `FINAL_REPORT` re-run - `PY_DERIVE` is served from cache.

### Visualise the pipeline

Nextflow can emit a DAG and an execution report alongside the run:

```bash
nextflow run main.nf \
  -with-dag flowchart.svg \
  -with-report report.html \
  -with-timeline timeline.html
```

- `flowchart.svg` - the workflow DAG (needs Graphviz / `dot`)
- `report.html` - per-task runtime, memory, and CPU usage
- `timeline.html` - Gantt-style view of when each task ran

Open the `.html` files in your browser.

---

## Snakemake

**Directory:** `demo/snakemake/`

The pipeline is defined in `Snakefile`. Each step is a `rule` block. Snakemake infers execution order by matching `output:` of one rule to `input:` of the next. Intermediate outputs go to `work/`, the final report to `results/`.

### Running

```bash
# Build the image and start the container (first run only)
docker compose up -d

# Open a shell inside the container
docker compose exec snakemake bash

# Run the pipeline
snakemake --cores 1
```

If you run Snakemake from outside `demo/snakemake/`, pass an explicit directory and Snakefile:

```bash
snakemake \
  --directory /workspace/contents/module_3_workflow_management/demo/snakemake \
  --snakefile /workspace/contents/module_3_workflow_management/demo/snakemake/Snakefile \
  --cores 1
```

Results are written to `results/4_report.html`.

### Re-running after a change

Snakemake tracks file modification timestamps. Just re-run — it automatically skips steps whose outputs are newer than their inputs:

```bash
snakemake --cores 1
```

To force a specific rule to re-run regardless of timestamps:

```bash
snakemake --cores 1 --forcerun r_summary
```

### Visualise the pipeline

Snakemake can emit the dependency graph as Graphviz `dot` format and an HTML execution report after a run:

```bash
# Rule-level DAG (one node per rule)
snakemake --rulegraph | dot -Tsvg > rulegraph.svg

# Full job DAG (one node per executed job, useful when wildcards fan out)
snakemake --dag | dot -Tsvg > dag.svg

# HTML report — run the pipeline first, then ask for the report
snakemake --cores 1
snakemake --report report.html
```

Open `rulegraph.svg`, `dag.svg`, or `report.html` in your browser. The report includes per-rule runtime, the DAG coloured by status, and any outputs annotated with `report()` in the Snakefile.

> Requires Graphviz (`dot`). The workshop image installs it for you.

### Your turn: write the dynamic report

In the Snakemake variant, the final pipeline step is **yours to write**.
`4_report.qmd` ships as a skeleton with "Your task" callouts. Apply what you
learned in the dynamic reporting hands-on ([R version](../../module_2_dynamic_reporting/demo_R/index.qmd),
[Python version](../../module_2_dynamic_reporting/demo_py/index.qmd)) to turn it
into a dynamic report that consumes the pipeline outputs:

1. **Setup**: load the libraries of your language of choice.
2. **Derived data**: read `work/1_derived.csv` and display it as a table with a caption.
3. **Summary statistics**: read `work/2_summary.txt` with code and print it.
4. **Processed data**: read `work/3_processed.csv` and plot `value_centered_z` over `year`, with a figure label and caption you can cross-reference.
5. **Bonus**: report a computed value (e.g., the number of rows) as inline code.

The workflow loop is the point of the exercise: after each edit to
`4_report.qmd`, re-run

```bash
snakemake --cores 1
```

and open `results/4_report.html`. Snakemake notices that only the report file
changed and re-runs just the `final_report` rule. The analysis steps are
skipped. If a code chunk errors, `quarto render` fails and Snakemake reports
the `final_report` rule as failed.

For contrast, the Nextflow variant ships a finished *static* report that embeds
the output files with `{{{< include >}}}` shortcodes and executes no code. Your
Snakemake report is *dynamic*: tables, figures, and inline numbers are computed
from the pipeline outputs at render time, so they can never go stale.

::: {.callout-tip collapse="true"}
#### Hint: which functions to reach for

| Task | R | Python |
|------|---|--------|
| Read a CSV | `read.csv()` | `pd.read_csv()` |
| Show a table | `kable()` from `kableExtra` | `itables.show()` |
| Read/print a text file | `readLines()` + `cat()` | `Path(...).read_text()` + `print()` |
| Plot | `ggplot()` + `geom_col()` | `sns.barplot()` |
| Inline code | `` `{{r}} nrow(df)` `` | `` `{{python}} len(df)` `` |

All libraries are pre-installed in the workshop container. Remember the chunk
options `#| label: fig-...` and `#| fig-cap:` for the figure, and that paths
are relative to the `demo/snakemake/` directory where the report is rendered.
:::

::: {.callout-tip collapse="true"}
#### Solution (R)

Replace the contents of `4_report.qmd` with:

````` markdown
---
title: "Module 3 Demo: Final Report"
format:
  html:
    toc: true
    embed-resources: true
execute:
  warning: false
---

This report is rendered as the final step of the Snakemake pipeline and reads
the artifacts produced by the earlier steps.

## Setup

```{r}
#| output: false
library("dplyr")      # data wrangling
library("ggplot2")    # plotting
library("kableExtra") # tables
```

## Derived data

```{r}
derived <- read.csv("work/1_derived.csv")
derived |>
  kable(caption = "Derived data: values centred by the Python step.")
```

## Summary statistics

```{r}
cat(readLines("work/2_summary.txt"), sep = "\n")
```

## Processed data

```{r}
#| label: fig-zscores
#| fig-cap: "Z-scores of the centred values by year."
#| fig-height: 4
processed <- read.csv("work/3_processed.csv")
processed |>
  ggplot(aes(x = factor(year), y = value_centered_z)) +
  geom_col() +
  labs(x = "Year", y = "Z-score") +
  theme_minimal()
```

@fig-zscores shows how far each year deviates from the mean.

## Bonus: inline statistics

The processed dataset contains `{r} nrow(processed)` rows covering the years
`{r} min(processed$year)` to `{r} max(processed$year)`.
`````
:::

::: {.callout-tip collapse="true"}
#### Solution (Python)

Replace the contents of `4_report.qmd` with:

````` markdown
---
title: "Module 3 Demo: Final Report"
format:
  html:
    toc: true
    embed-resources: true
execute:
  warning: false
---

This report is rendered as the final step of the Snakemake pipeline and reads
the artifacts produced by the earlier steps.

## Setup

```{python}
#| output: false
import pandas as pd              # data wrangling
import itables                   # nicer tables
import matplotlib.pyplot as plt  # plotting
import seaborn as sns            # nicer plots
from pathlib import Path

sns.set()  # activate seaborn defaults
```

## Derived data

```{python}
derived = pd.read_csv("work/1_derived.csv")
itables.show(derived, caption="Derived data: values centred by the Python step.")
```

## Summary statistics

```{python}
print(Path("work/2_summary.txt").read_text())
```

## Processed data

```{python}
#| label: fig-zscores
#| fig-cap: "Z-scores of the centred values by year."
processed = pd.read_csv("work/3_processed.csv")
ax = sns.barplot(processed, x="year", y="value_centered_z")
ax.set(xlabel="Year", ylabel="Z-score")
plt.tight_layout()
```

@fig-zscores shows how far each year deviates from the mean.

## Bonus: inline statistics

The processed dataset contains `{python} len(processed)` rows covering the years
`{python} int(processed["year"].min())` to `{python} int(processed["year"].max())`.
`````
:::

---

## Key differences

| | Nextflow | Snakemake |
|---|---|---|
| Pipeline file | `main.nf` | `Snakefile` |
| Step isolation | Each task runs in its own `work/<hash>/` subdirectory | Steps run in the project directory |
| Intermediate outputs | Staged in `work/`, copied to `results/` via `publishDir` | Written directly to `work/` |
| Cache mechanism | Content hashes (task inputs + code) | File modification timestamps |
| Re-run only changed steps | `nextflow run main.nf -resume` | `snakemake --cores 1` (automatic) |
| Force re-run one step | Not directly — delete its `work/` entry or use `-resume` after touching the script | `--forcerun <rule_name>` |
